---
title: "01 — Modeling Setup (PostgreSQL → R)"
author: "Waldo Ketonou"
output: html_document
---

# Introduction
This notebook begins the R modeling workflow. All data cleaning, joins, feature engineering, and EDA were completed in PostgreSQL. The final modeling‑ready table is `modeling_clean`. This notebook connects to PostgreSQL, loads the prepared dataset, performs modeling‑readiness checks, converts variables, and creates the time‑based train/test split used in downstream modeling scripts.

# Load Libraries
```{r}
library(DBI)
library(RPostgres)
library(dplyr)
library(lubridate)
```
# Connect to PostgreSQL
```{r}
con <- dbConnect(
  RPostgres::Postgres(),
  dbname   = "walmart_sales",
  host     = "localhost",
  port     = 5432,
  user     = "postgres",
  password = "mypasswordtopostegresql"
)
```
# Load modeling_clean
```{r}
df <- dbReadTable(con, "modeling_clean")
glimpse(df)
```
# Basic Modeling Checks
```{r}
summary(df)
sapply(df, function(x) length(unique(x)))
```
# Prepare Variable
```{r}
   df <- df %>%
  mutate(
    date = as.Date(date),
    store_type = as.factor(store_type),
    feature_is_holiday = as.factor(feature_is_holiday)
  )

names(df)[names(df) == "weekly_sales"] <- "actual_sales"
```
# Time-Based Train/Test Split
```{r}
train_df <- df %>% filter(year < 2012)
test_df  <- df %>% filter(year == 2012)

range(train_df$date)
range(test_df$date)
```
# Save Local Copy for Modeling
```{r}
dir.create("data", showWarnings = FALSE)
dir.create("data/cleaned", showWarnings = FALSE)

write.csv(df, "data/cleaned/modeling_clean.csv", row.names = FALSE)
```

 
