---
title: "02 – Model Training Notebook"
author: "Waldo Ketonou | WaldoSphere Group LLC"
output: html_document
---

# Walmart Weekly Sales Forecasting  
## Notebook 02 – Model Training  

### Introduction  
This notebook documents the full modeling workflow for the Walmart Weekly Sales Forecasting project.  
It mirrors the production script `07_model_training.R` and:

- Trains a baseline linear regression model using unified predictors  
- Runs Random Forest scaling experiments (SMALL1: 10k, SMALL2: 100k, SMALL3: 200k, MEDIUM: 500k, LARGE: 1M)  
- Saves all model objects, predictions, and evaluation metrics for downstream analysis, visualization, and scenario simulation.

---

### 1. Environment Setup

```{r setup, message=FALSE, warning=FALSE}
rm(list = ls())
gc()
cat("\014")
if (!is.null(dev.list())) dev.off()

library(dplyr)
library(readr)
library(ranger)

cat("Environment cleared. Ready to load train/test CSVs.\n")
```
### 2. Baseline Linear Regression Model
```{r}
cat("\n===== Linear Regression: BASELINE MODEL =====\n")

train_clean <- read_csv("outputs/prepared_data/train_df.csv")
test_df     <- read_csv("outputs/prepared_data/test_df.csv")
test_clean  <- test_df %>% filter(complete.cases(.))

smape <- function(actual, predicted) {
  mean(2 * abs(predicted - actual) / (abs(actual) + abs(predicted))) * 100
}

lm_formula <- actual_sales ~ 
  lag_4_weeks + rolling_4_week + rolling_12_week +
  feature_is_holiday + temperature + fuel_price + cpi + unemployment +
  store_type + store_size + week + month + year

lm_model <- lm(lm_formula, data = train_clean)
lm_pred  <- predict(lm_model, newdata = test_clean)

eval_idx    <- !is.na(lm_pred)
eval_pred   <- lm_pred[eval_idx]
eval_actual <- test_clean$actual_sales[eval_idx]

rmse_lm  <- sqrt(mean((eval_pred - eval_actual)^2))
mae_lm   <- mean(abs(eval_pred - eval_actual))
mape_lm  <- mean(abs((eval_pred - eval_actual) / eval_actual)) * 100
smape_lm <- smape(eval_actual, eval_pred)

dir.create("outputs/models", recursive = TRUE, showWarnings = FALSE)
dir.create("outputs/predictions", recursive = TRUE, showWarnings = FALSE)
dir.create("outputs/evaluation", recursive = TRUE, showWarnings = FALSE)

saveRDS(lm_model, "outputs/models/lm_baseline_model.rds")

write_csv(
  tibble(actual = eval_actual, predicted = eval_pred),
  "outputs/predictions/lm_baseline_predictions.csv"
)

write_csv(
  tibble(rmse = rmse_lm, mae = mae_lm, mape = mape_lm, smape = smape_lm),
  "outputs/evaluation/lm_baseline_metrics.csv"
)

cat("RMSE :", rmse_lm, "\n")
cat("MAE  :", mae_lm, "\n")
cat("MAPE :", mape_lm, "%\n")
cat("SMAPE:", smape_lm, "%\n")

rm(list = ls())
gc()
```
### 3. Random Forest Scaling Eexperiments 
We train RF models at sample size: 10k → 100k → 200k → 500k → 1M
## 3A. SMALL1 — 10,000 rows
```{r}
cat("\n===== Random Forest: SMALL1 (10k rows) =====\n")

train_clean <- read_csv("outputs/prepared_data/train_df.csv")
test_df     <- read_csv("outputs/prepared_data/test_df.csv")
test_clean  <- test_df %>% filter(complete.cases(.))

smape <- function(actual, predicted) {
  mean(2 * abs(predicted - actual) / (abs(actual) + abs(predicted))) * 100
}

set.seed(123)
sampled_train <- train_clean %>% filter(complete.cases(.)) %>% sample_n(10000)

rf_small1_model <- ranger(
  formula = actual_sales ~ lag_4_weeks + rolling_4_week + rolling_12_week +
    feature_is_holiday + temperature + fuel_price + cpi + unemployment +
    store_type + store_size + week + month + year,
  data = sampled_train,
  num.trees = 200,
  write.forest = TRUE,
  num.threads = 1,
  importance = "impurity"
)

rf_small1_pred   <- predict(rf_small1_model, data = test_clean)$predictions
rf_small1_actual <- test_clean$actual_sales

rf_small1_rmse  <- sqrt(mean((rf_small1_pred - rf_small1_actual)^2))
rf_small1_mae   <- mean(abs(rf_small1_pred - rf_small1_actual))
rf_small1_mape  <- mean(abs((rf_small1_pred - rf_small1_actual) / rf_small1_actual)) * 100
rf_small1_smape <- smape(rf_small1_actual, rf_small1_pred)

saveRDS(rf_small1_model, "outputs/models/rf_small1_model.rds")

write_csv(
  tibble(actual = rf_small1_actual, predicted = rf_small1_pred),
  "outputs/predictions/rf_small1_predictions.csv"
)

write_csv(
  tibble(rmse = rf_small1_rmse, mae = rf_small1_mae, mape = rf_small1_mape, smape = rf_small1_smape),
  "outputs/evaluation/rf_small1_metrics.csv"
)

cat("RMSE :", rf_small1_rmse, "\n")
cat("MAE  :", rf_small1_mae, "\n")
cat("MAPE :", rf_small1_mape, "%\n")
cat("SMAPE:", rf_small1_smape, "%\n")

rm(list = ls())
gc()
```
## 3B. SMALL2 — 100,000 rows
```{r}
cat("\n===== Random Forest: SMALL2 (100k rows) =====\n")

train_clean <- read_csv("outputs/prepared_data/train_df.csv")
test_df     <- read_csv("outputs/prepared_data/test_df.csv")
test_clean  <- test_df %>% filter(complete.cases(.))

smape <- function(actual, predicted) {
  mean(2 * abs(predicted - actual) / (abs(actual) + abs(predicted))) * 100
}

set.seed(123)
sampled_train <- train_clean %>% filter(complete.cases(.)) %>% sample_n(100000)

rf_small2_model <- ranger(
  formula = actual_sales ~ lag_4_weeks + rolling_4_week + rolling_12_week +
    feature_is_holiday + temperature + fuel_price + cpi + unemployment +
    store_type + store_size + week + month + year,
  data = sampled_train,
  num.trees = 200,
  write.forest = TRUE,
  num.threads = 1,
  importance = "impurity"
)

rf_small2_pred   <- predict(rf_small2_model, data = test_clean)$predictions
rf_small2_actual <- test_clean$actual_sales

rf_small2_rmse  <- sqrt(mean((rf_small2_pred - rf_small2_actual)^2))
rf_small2_mae   <- mean(abs(rf_small2_pred - rf_small2_actual))
rf_small2_mape  <- mean(abs((rf_small2_pred - rf_small2_actual) / rf_small2_actual)) * 100
rf_small2_smape <- smape(rf_small2_actual, rf_small2_pred)

saveRDS(rf_small2_model, "outputs/models/rf_small2_model.rds")

write_csv(
  tibble(actual = rf_small2_actual, predicted = rf_small2_pred),
  "outputs/predictions/rf_small2_predictions.csv"
)

write_csv(
  tibble(rmse = rf_small2_rmse, mae = rf_small2_mae, mape = rf_small2_mape, smape = rf_small2_smape),
  "outputs/evaluation/rf_small2_metrics.csv"
)

cat("RMSE :", rf_small2_rmse, "\n")
cat("MAE  :", rf_small2_mae, "\n")
cat("MAPE :", rf_small2_mape, "%\n")
cat("SMAPE:", rf_small2_smape, "%\n")

rm(list = ls())
gc()
```
## 3C. SMALL3 — 200,000 rows
```{r}
cat("\n===== Random Forest: SMALL3 (200k rows) =====\n")

train_clean <- read_csv("outputs/prepared_data/train_df.csv")
test_df     <- read_csv("outputs/prepared_data/test_df.csv")
test_clean  <- test_df %>% filter(complete.cases(.))

smape <- function(actual, predicted) {
  mean(2 * abs(predicted - actual) / (abs(actual) + abs(predicted))) * 100
}

set.seed(123)
sampled_train <- train_clean %>% filter(complete.cases(.)) %>% sample_n(200000)

rf_small3_model <- ranger(
  formula = actual_sales ~ lag_4_weeks + rolling_4_week + rolling_12_week +
    feature_is_holiday + temperature + fuel_price + cpi + unemployment +
    store_type + store_size + week + month + year,
  data = sampled_train,
  num.trees = 200,
  write.forest = TRUE,
  num.threads = 1,
  importance = "impurity"
)

rf_small3_pred   <- predict(rf_small3_model, data = test_clean)$predictions
rf_small3_actual <- test_clean$actual_sales

rf_small3_rmse  <- sqrt(mean((rf_small3_pred - rf_small3_actual)^2))
rf_small3_mae   <- mean(abs(rf_small3_pred - rf_small3_actual))
rf_small3_mape  <- mean(abs((rf_small3_pred - rf_small3_actual) / rf_small3_actual)) * 100
rf_small3_smape <- smape(rf_small3_actual, rf_small3_pred)

saveRDS(rf_small3_model, "outputs/models/rf_small3_model.rds")

write_csv(
  tibble(actual = rf_small3_actual, predicted = rf_small3_pred),
  "outputs/predictions/rf_small3_predictions.csv"
)

write_csv(
  tibble(rmse = rf_small3_rmse, mae = rf_small3_mae, mape = rf_small3_mape, smape = rf_small3_smape),
  "outputs/evaluation/rf_small3_metrics.csv"
)

cat("RMSE :", rf_small3_rmse, "\n")
cat("MAE  :", rf_small3_mae, "\n")
cat("MAPE :", rf_small3_mape, "%\n")
cat("SMAPE:", rf_small3_smape, "%\n")

rm(list = ls())
gc()
```
## 3D. MEDIUM — 500,000 rows
```{r}
cat("\n===== Random Forest: MEDIUM (500k rows) =====\n")

train_clean <- read_csv("outputs/prepared_data/train_df.csv")
test_df     <- read_csv("outputs/prepared_data/test_df.csv")
test_clean  <- test_df %>% filter(complete.cases(.))

smape <- function(actual, predicted) {
  mean(2 * abs(predicted - actual) / (abs(actual) + abs(predicted))) * 100
}

set.seed(123)
sampled_train <- train_clean %>% filter(complete.cases(.)) %>% sample_n(500000)

rf_medium_model <- ranger(
  formula = actual_sales ~ lag_4_weeks + rolling_4_week + rolling_12_week +
    feature_is_holiday + temperature + fuel_price + cpi + unemployment +
    store_type + store_size + week + month + year,
  data = sampled_train,
  num.trees = 200,
  write.forest = TRUE,
  num.threads = 1,
  importance = "impurity"
)

rf_medium_pred   <- predict(rf_medium_model, data = test_clean)$predictions
rf_medium_actual <- test_clean$actual_sales

rf_medium_rmse  <- sqrt(mean((rf_medium_pred - rf_medium_actual)^2))
rf_medium_mae   <- mean(abs(rf_medium_pred - rf_medium_actual))
rf_medium_mape  <- mean(abs((rf_medium_pred - rf_medium_actual) / rf_medium_actual)) * 100
rf_medium_smape <- smape(rf_medium_actual, rf_medium_pred)

saveRDS(rf_medium_model, "outputs/models/rf_medium_model.rds")

write_csv(
  tibble(actual = rf_medium_actual, predicted = rf_medium_pred),
  "outputs/predictions/rf_medium_predictions.csv"
)

write_csv(
  tibble(rmse = rf_medium_rmse, mae = rf_medium_mae, mape = rf_medium_mape, smape = rf_medium_smape),
  "outputs/evaluation/rf_medium_metrics.csv"
)

cat("RMSE :", rf_medium_rmse, "\n")
cat("MAE  :", rf_medium_mae, "\n")
cat("MAPE :", rf_medium_mape, "%\n")
cat("SMAPE:", rf_medium_smape, "%\n")

rm(list = ls())
gc()
```
## 3E. LARGE — 1,000,000 rows 
```{r}
cat("\n===== Random Forest: LARGE (1M rows) =====\n")

train_clean <- read_csv("outputs/prepared_data/train_df.csv")
test_df     <- read_csv("outputs/prepared_data/test_df.csv")
test_clean  <- test_df %>% filter(complete.cases(.))

smape <- function(actual, predicted) {
  mean(2 * abs(predicted - actual) / (abs(actual) + abs(predicted))) * 100
}

set.seed(123)
sampled_train <- train_clean %>% filter(complete.cases(.)) %>% sample_n(1000000)

rf_large_model <- ranger(
  formula = actual_sales ~ lag_4_weeks + rolling_4_week + rolling_12_week +
    feature_is_holiday + temperature + fuel_price + cpi + unemployment +
    store_type + store_size + week + month + year,
  data = sampled_train,
  num.trees = 200,
  write.forest = TRUE,
  num.threads = 1,
  importance = "impurity"
)

rf_large_pred   <- predict(rf_large_model, data = test_clean)$predictions
rf_large_actual <- test_clean$actual_sales

rf_large_rmse  <- sqrt(mean((rf_large_pred - rf_large_actual)^2))
rf_large_mae   <- mean(abs(rf_large_pred - rf_large_actual))
rf_large_mape  <- mean(abs((rf_large_pred - rf_large_actual) / rf_large_actual)) * 100
rf_large_smape <- smape(rf_large_actual, rf_large_pred)

saveRDS(rf_large_model, "outputs/models/rf_large_model.rds")

write_csv(
  tibble(actual = rf_large_actual, predicted = rf_large_pred),
  "outputs/predictions/rf_large_predictions.csv"
)

write_csv(
  tibble(rmse = rf_large_rmse, mae = rf_large_mae, mape = rf_large_mape, smape = rf_large_smape),
  "outputs/evaluation/rf_large_metrics.csv"
)

cat("RMSE :", rf_large_rmse, "\n")
cat("MAE  :", rf_large_mae, "\n")
cat("MAPE :", rf_large_mape, "%\n")
cat("SMAPE:", rf_large_smape, "%\n")

rm(list = ls())
gc()
```
### 4. Diagnostic Visualizations
```{r}
if (!dir.exists("visuals")) dir.create("visuals")

## Linear Model: Predicted vs Actual
png("visuals/lm_pred_vs_actual.png", width = 1200, height = 800)
plot(eval_actual, eval_pred,
     main = "Linear Model: Predicted vs Actual",
     xlab = "Actual Sales",
     ylab = "Predicted Sales",
     pch = 16, col = rgb(0, 0, 1, 0.4))
abline(0, 1, col = "red", lwd = 2)
dev.off()

## Linear Model: Residuals
png("visuals/lm_residuals.png", width = 1200, height = 800)
plot(eval_pred - eval_actual,
     main = "Linear Model Residuals",
     xlab = "Index",
     ylab = "Residual",
     pch = 16, col = rgb(1, 0, 0, 0.4))
abline(h = 0, col = "blue", lwd = 2)
dev.off()
```
### 5. Conclusion
All models have been trained and saved.
