---
title: "03 – Model Comparison and Visualization"
author: "Waldo Ketonou | WaldoSphere Group LLC"
output: html_document
---

# Walmart Weekly Sales Forecasting  
## Notebook 03 – Model Comparison & Visualization  

### Introduction  
This notebook compares the performance of all trained models using the evaluation metrics defined in the project pipeline.  
It loads predictions generated in Notebook 02 / Script 07, computes RMSE, MAE, MAPE, and SMAPE, identifies the best-performing model, and generates visualizations including:

- Actual vs Predicted (best model)
- Residual distribution
- Predicted vs Actual scatter

All outputs are saved to the `/visuals` directory for use in the final case study.

---

# 1. Load Libraries

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(ggplot2)
library(readr)

cat("\n===== Model Comparison and Visualization =====\n")
```
# 2. Load Evaluation Data
We load the cleaned test set to obtain the actual sales values.

```{r}
test_df <- read_csv("outputs/prepared_data/test_df.csv")
test_clean <- test_df %>% filter(complete.cases(.))
eval_actual <- test_clean$actual_sales
```
# 3. Load Predictions
These prediction files were generated in Notebook 02 / Script 07.
```{r}
lm_pred        <- read_csv("outputs/predictions/lm_baseline_predictions.csv")$predicted
rf_small1_pred <- read_csv("outputs/predictions/rf_small1_predictions.csv")$predicted
rf_small2_pred <- read_csv("outputs/predictions/rf_small2_predictions.csv")$predicted
rf_small3_pred <- read_csv("outputs/predictions/rf_small3_predictions.csv")$predicted
rf_medium_pred <- read_csv("outputs/predictions/rf_medium_predictions.csv")$predicted
rf_large_pred  <- read_csv("outputs/predictions/rf_large_predictions.csv")$predicted
```
# 4. Define Evaluation Metrics
```{r}
rmse <- function(a, p) sqrt(mean((p - a)^2))
mae  <- function(a, p) mean(abs(p - a))
mape <- function(a, p) mean(abs((p - a) / a)) * 100
smape <- function(a, p) mean(2 * abs(p - a) / (abs(a) + abs(p))) * 100
```
# 5. Build Comparison Table
We compute all metrics for each model and assemble them into a single tibble.
```{r}
compare_tbl <- tibble(
  model = c(
    "Linear Regression",
    "RF Small1 (10k)",
    "RF Small2 (100k)",
    "RF Small3 (200k)",
    "RF Medium (500k)",
    "RF Large (1M)"
  ),
  RMSE  = c(
    rmse(eval_actual, lm_pred),
    rmse(eval_actual, rf_small1_pred),
    rmse(eval_actual, rf_small2_pred),
    rmse(eval_actual, rf_small3_pred),
    rmse(eval_actual, rf_medium_pred),
    rmse(eval_actual, rf_large_pred)
  ),
  MAE   = c(
    mae(eval_actual, lm_pred),
    mae(eval_actual, rf_small1_pred),
    mae(eval_actual, rf_small2_pred),
    mae(eval_actual, rf_small3_pred),
    mae(eval_actual, rf_medium_pred),
    mae(eval_actual, rf_large_pred)
  ),
  MAPE  = c(
    mape(eval_actual, lm_pred),
    mape(eval_actual, rf_small1_pred),
    mape(eval_actual, rf_small2_pred),
    mape(eval_actual, rf_small3_pred),
    mape(eval_actual, rf_medium_pred),
    mape(eval_actual, rf_large_pred)
  ),
  SMAPE = c(
    smape(eval_actual, lm_pred),
    smape(eval_actual, rf_small1_pred),
    smape(eval_actual, rf_small2_pred),
    smape(eval_actual, rf_small3_pred),
    smape(eval_actual, rf_medium_pred),
    smape(eval_actual, rf_large_pred)
  )
)

compare_tbl
```
# 6. Identify Best Model
We select the model with the lowest RMSE.
```{r}
best_model_name <- compare_tbl$model[which.min(compare_tbl$RMSE)]

best_pred <- dplyr::case_when(
  best_model_name == "Linear Regression" ~ lm_pred,
  best_model_name == "RF Small1 (10k)"   ~ rf_small1_pred,
  best_model_name == "RF Small2 (100k)"  ~ rf_small2_pred,
  best_model_name == "RF Small3 (200k)"  ~ rf_small3_pred,
  best_model_name == "RF Medium (500k)"  ~ rf_medium_pred,
  TRUE                                   ~ rf_large_pred
)

cat("\nBest Model Identified:", best_model_name, "\n")
```
# 7. Prepare Visualization Data
We build a tidy dataframe for plotting.
```{r}
viz_df <- tibble(
  date      = test_clean$date,
  actual    = eval_actual,
  predicted = best_pred,
  residuals = eval_actual - best_pred
)

dir.create("visuals", recursive = TRUE, showWarnings = FALSE)
```
# 8. Visualization — Actual vs Predicted
A time‑series comparison of actual vs predicted sales.
```{r}
p1 <- ggplot(viz_df, aes(x = date)) +
  geom_line(aes(y = actual, color = "Actual"), linewidth = 1.2) +
  geom_line(aes(y = predicted, color = "Predicted"), linewidth = 1.2) +
  labs(
    title = paste("Actual vs Predicted Sales —", best_model_name),
    x = "Date", y = "Sales"
  ) +
  scale_color_manual(values = c("Actual" = "black", "Predicted" = "blue")) +
  theme_minimal(base_size = 14)

ggsave("visuals/best_model_actual_vs_predicted.png", p1, width = 10, height = 6)
```
# 9. Visualization — Residual Distribution
This helps assess model bias and error spread.
```{r}
p2 <- ggplot(viz_df, aes(x = residuals)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  labs(
    title = paste("Residual Distribution —", best_model_name),
    x = "Residuals", y = "Count"
  ) +
  theme_minimal(base_size = 14)

ggsave("visuals/best_model_residual_distribution.png", p2, width = 8, height = 5)
```
# 0. Visualization — Predicted vs Actual Scatter
A diagnostic scatterplot with a 45° reference line.
```{r}
p3 <- ggplot(viz_df, aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.4, color = "darkgreen") +
  geom_abline(slope = 1, intercept = 0, color = "red", linewidth = 1.2) +
  labs(
    title = paste("Predicted vs Actual Scatter —", best_model_name),
    x = "Actual Sales", y = "Predicted Sales"
  ) +
  theme_minimal(base_size = 14)

ggsave("visuals/best_model_pred_vs_actual_scatter.png", p3, width = 8, height = 5)

cat("\nSaved visualizations to /visuals/\n")
```
# 11. Conclusion
This notebook successfully:

Compared all trained models

Identified the best-performing model

Generated three key diagnostic visualizations

Saved all outputs for use in the final case study
