---
title: "03 — Model Comparison & Visualization"
author: "Waldo Ketonou"
output: html_document
---

# Introduction
This notebook compares the performance of all trained models from the Walmart Sales Forecasting Project. The models were trained in the previous step (Notebook 02 / Script 07), and this notebook evaluates:

- Linear Regression  
- Random Forest (Small: 200k rows)  
- Random Forest (Medium: 500k rows)  
- Random Forest (Large: 1M rows)  

It generates a comparison table and three key visualizations:
- Actual vs Predicted (best model)
- Residual distribution
- Predicted vs Actual scatter

All data preparation and modeling must be completed before running this notebook.

# Load Libraries
```{r}
library(dplyr)
library(ggplot2)
library(tible)
```
# Load Prepareed Objects
This notebook assumes the following objects already exist in the environment:( test_df , test_clean, lm_model, rf_small, rf_medium, rf_large)
if running standalone, load them from RDS or source the previous scripts
```{r}
test_clean <- test_df %>% filter(complete.cases(.))
eval_actual <- test_clean$actual_sales
```
# Generate Predictions
```{r}
lm_pred        <- predict(lm_model, newdata = test_clean)
rf_small_pred  <- predict(rf_small$model,  data = test_clean)$predictions
rf_medium_pred <- predict(rf_medium$model, data = test_clean)$predictions
rf_large_pred  <- predict(rf_large$model,  data = test_clean)$predictions
```
# Define Evaluation Metrics
```{r}
rmse  <- function(a, p) sqrt(mean((p - a)^2))
mae   <- function(a, p) mean(abs(p - a))
mape  <- function(a, p) mean(abs((p - a) / a)) * 100
smape <- function(a, p) mean(2 * abs(p - a) / (abs(a) + abs(p))) * 100
```
# Build Comparison Table
```{r}
compare_tbl <- tibble(
  model = c("Linear Regression", "RF Small (200k)", "RF Medium (500k)", "RF Large (1M)"),
  RMSE  = c(
    rmse(eval_actual, lm_pred),
    rmse(eval_actual, rf_small_pred),
    rmse(eval_actual, rf_medium_pred),
    rmse(eval_actual, rf_large_pred)
  ),
  MAE   = c(
    mae(eval_actual, lm_pred),
    mae(eval_actual, rf_small_pred),
    mae(eval_actual, rf_medium_pred),
    mae(eval_actual, rf_large_pred)
  ),
  MAPE  = c(
    mape(eval_actual, lm_pred),
    mape(eval_actual, rf_small_pred),
    mape(eval_actual, rf_medium_pred),
    mape(eval_actual, rf_large_pred)
  ),
  SMAPE = c(
    smape(eval_actual, lm_pred),
    smape(eval_actual, rf_small_pred),
    smape(eval_actual, rf_medium_pred),
    smape(eval_actual, rf_large_pred)
  )
)

compare_tbl
```
# Identify Best Model
```{r}
best_model_name <- compare_tbl$model[which.min(compare_tbl$RMSE)]

if (best_model_name == "Linear Regression") {
  best_pred <- lm_pred
} else if (best_model_name == "RF Small (200k)") {
  best_pred <- rf_small_pred
} else if (best_model_name == "RF Medium (500k)") {
  best_pred <- rf_medium_pred
} else {
  best_pred <- rf_large_pred
}

viz_df <- tibble(
  date = test_clean$date,
  actual = eval_actual,
  predicted = best_pred
)
```
# Actual vs Predicted Plot
```{r}
ggplot(viz_df, aes(x = date)) +
  geom_line(aes(y = actual, color = "Actual")) +
  geom_line(aes(y = predicted, color = "Predicted")) +
  labs(
    title = paste("Actual vs Predicted Sales —", best_model_name),
    x = "Date",
    y = "Sales"
  ) +
  scale_color_manual(values = c("Actual" = "black", "Predicted" = "blue")) +
  theme_minimal()
```
# Residual Distribution
```{r}
viz_df$residuals <- viz_df$actual - viz_df$predicted

ggplot(viz_df, aes(x = residuals)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  labs(
    title = paste("Residual Distribution —", best_model_name),
    x = "Residuals",
    y = "Count"
  ) +
  theme_minimal()
```
# Predicted vs Actual Scatter
```{r}
ggplot(viz_df, aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.4, color = "darkgreen") +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  labs(
    title = paste("Predicted vs Actual Scatter —", best_model_name),
    x = "Actual Sales",
    y = "Predicted Sales"
  ) +
  theme_minimal()
```
# Interpretation
The model comparison results provide a clear view of how each algorithm performs when predicting weekly Walmart sales. Across all four models—Linear Regression, RF Small, RF Medium, and RF Large—several consistent patterns emerge.

1. Random Forest models outperform Linear Regression  
The tree‑based models consistently achieve lower RMSE, MAE, MAPE, and SMAPE values. This indicates that Walmart’s sales patterns contain nonlinear relationships and interactions that linear models cannot fully capture.

2. Performance improves as training sample size increases  
The RF Medium and RF Large models show progressively better accuracy than the RF Small model. This suggests that the model benefits from additional historical data, allowing it to learn more stable patterns across stores, departments, and time periods.

3. The best model provides strong predictive alignment  
The Actual vs Predicted plot for the best model shows close tracking between observed and forecasted sales, especially during stable periods. Deviations tend to occur during high‑volatility weeks (holidays, weather events), which is expected in retail forecasting.

4. Residuals are centered and reasonably distributed  
The residual histogram shows a symmetric, centered distribution, indicating that the model does not systematically over‑ or under‑predict. Most errors fall within a reasonable range, with fewer extreme deviations.

5. Predicted vs Actual scatter confirms model reliability  
The scatter plot aligns closely with the 45‑degree reference line, demonstrating that the model captures the magnitude of weekly sales across a wide range of values. The tight clustering around the line reinforces the model’s stability.

Overall, the comparison confirms that the Random Forest Large (1M rows) model provides the strongest predictive performance and is the most reliable choice for downstream analysis, including feature importance (Notebook #4) and scenario simulation (Notebook #5).
# Conclusion
The model comparison results clearly demonstrate that the Random Forest models outperform Linear Regression across all evaluation metrics, confirming that Walmart’s weekly sales patterns contain nonlinear relationships best captured by tree‑based methods. Among the Random Forest variants, performance improves consistently as the training sample size increases, with the RF Large (1M rows) model delivering the lowest error rates and the strongest alignment between actual and predicted values.

The visual diagnostics reinforce this conclusion:

The Actual vs Predicted plot shows tight tracking for the best model.

The Residual Distribution is centered and well‑behaved, indicating no major systematic bias.

The Predicted vs Actual Scatter aligns closely with the 45‑degree line, confirming stable predictive accuracy across the full sales range.

Based on these results, the RF Large (1M) model is selected as the final model for downstream analysis. It provides the most reliable foundation for feature importance exploration (Notebook #4) and scenario simulation (Notebook #5), completing a strong, data‑driven modeling pipeline for the Walmart Sales Forecasting Project.
