---
title: "03 — Model Comparison & Visualization"
author: "Waldo Ketonou"
output: html_document
---

# Introduction
This notebook compares the performance of all trained models from the Walmart Sales Forecasting Project. The models were trained in the previous step (Notebook 02 / Script 07), and this notebook evaluates:

- Linear Regression  
- Random Forest (Small: 200k rows)  
- Random Forest (Medium: 500k rows)  
- Random Forest (Large: 1M rows)  

It generates a comparison table and three key visualizations:
- Actual vs Predicted (best model)
- Residual distribution
- Predicted vs Actual scatter

All data preparation and modeling must be completed before running this notebook.

# Load Libraries
```{r}
library(dplyr)
library(ggplot2)
# Load Prepareed Objects
This notebook assumes the following objects already exist in the environment:( test_df , test_clean, lm_model, rf_small, rf_medium, rf_large)
if running standalone, load them from RDS or source the previous scripts
```{r}
test_clean <- test_df %>% filter(complete.cases(.))
eval_actual <- test_clean$actual_sales
# Generate Predictions
```{r}
lm_pred        <- predict(lm_model, newdata = test_clean)
rf_small_pred  <- predict(rf_small$model,  data = test_clean)$predictions
rf_medium_pred <- predict(rf_medium$model, data = test_clean)$predictions
rf_large_pred  <- predict(rf_large$model,  data = test_clean)$predictions
# Define Evaluation Metrics
```{r}
rmse  <- function(a, p) sqrt(mean((p - a)^2))
mae   <- function(a, p) mean(abs(p - a))
mape  <- function(a, p) mean(abs((p - a) / a)) * 100
smape <- function(a, p) mean(2 * abs(p - a) / (abs(a) + abs(p))) * 100
# Build Comparison Table
```{r}
compare_tbl <- tibble(
  model = c("Linear Regression", "RF Small (200k)", "RF Medium (500k)", "RF Large (1M)"),
  RMSE  = c(
    rmse(eval_actual, lm_pred),
    rmse(eval_actual, rf_small_pred),
    rmse(eval_actual, rf_medium_pred),
    rmse(eval_actual, rf_large_pred)
  ),
  MAE   = c(
    mae(eval_actual, lm_pred),
    mae(eval_actual, rf_small_pred),
    mae(eval_actual, rf_medium_pred),
    mae(eval_actual, rf_large_pred)
  ),
  MAPE  = c(
    mape(eval_actual, lm_pred),
    mape(eval_actual, rf_small_pred),
    mape(eval_actual, rf_medium_pred),
    mape(eval_actual, rf_large_pred)
  ),
  SMAPE = c(
    smape(eval_actual, lm_pred),
    smape(eval_actual, rf_small_pred),
    smape(eval_actual, rf_medium_pred),
    smape(eval_actual, rf_large_pred)
  )
)

compare_tbl
# Identify Best Model
```{r}
best_model_name <- compare_tbl$model[which.min(compare_tbl$RMSE)]

if (best_model_name == "Linear Regression") {
  best_pred <- lm_pred
} else if (best_model_name == "RF Small (200k)") {
  best_pred <- rf_small_pred
} else if (best_model_name == "RF Medium (500k)") {
  best_pred <- rf_medium_pred
} else {
  best_pred <- rf_large_pred
}

viz_df <- tibble(
  date = test_clean$date,
  actual = eval_actual,
  predicted = best_pred
)
# Actual vs Predicted Plot
```{r}
ggplot(viz_df, aes(x = date)) +
  geom_line(aes(y = actual, color = "Actual")) +
  geom_line(aes(y = predicted, color = "Predicted")) +
  labs(
    title = paste("Actual vs Predicted Sales —", best_model_name),
    x = "Date",
    y = "Sales"
  ) +
  scale_color_manual(values = c("Actual" = "black", "Predicted" = "blue")) +
  theme_minimal()
# Residual Distribution
```{r}
viz_df$residuals <- viz_df$actual - viz_df$predicted

ggplot(viz_df, aes(x = residuals)) +
  geom_histogram(bins = 50, fill = "steelblue", color = "white") +
  labs(
    title = paste("Residual Distribution —", best_model_name),
    x = "Residuals",
    y = "Count"
  ) +
  theme_minimal()
# Predicted vs Actual Scatter
```{r}
ggplot(viz_df, aes(x = actual, y = predicted)) +
  geom_point(alpha = 0.4, color = "darkgreen") +
  geom_abline(slope = 1, intercept = 0, color = "red") +
  labs(
    title = paste("Predicted vs Actual Scatter —", best_model_name),
    x = "Actual Sales",
    y = "Predicted Sales"
  ) +
  theme_minimal()
