---
title: "04 — Feature Importance & Model Insights"
author: "Waldo Ketonou"
output: html_document
---
# Introduction
This notebook analyzes feature importance for the Random Forest models trained in the Walmart Sales Forecasting Project.
Models included:
RF Small (200k rows)
RF Medium (500k rows)
RF Large (1M rows)
Goals:
Identify which features most strongly drive weekly sales
Compare importance patterns across different sample sizes
Provide model interpretation for the final case study and dashboard
All data preparation and model training must be completed before running this notebook.
# Load Librarie
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(tibble)
library(knitr)
```
# Load Trained Models
This notebook assumes the Random Forest models have already been traine and saved as RDS in the outputs/models/ directory.
If you are running from a different working directory, adjust the paths accordingly.
```{r}
rf_small_path  <- "../outputs/models/rf_small.rds"
rf_medium_path <- "../outputs/models/rf_medium.rds"
rf_large_path  <- "../outputs/models/rf_large.rds"

rf_small  <- readRDS(rf_small_path)
rf_medium <- readRDS(rf_medium_path)
rf_large  <- readRDS(rf_large_path)
```
# Extract Variable Importance
I define a helper function to extract variable importance from each Random Forest model.
```{r}
get_importance <- function(model, model_name) {
  
  if (is.null(model$variable.importance)) {
    stop(paste("Variable importance not found for model:", model_name))
  }
  
  tibble(
    feature    = names(model$variable.importance),
    importance = as.numeric(model$variable.importance),
    model      = model_name
  )
}

imp_small  <- get_importance(rf_small,  "RF Small (200k)")
imp_medium <- get_importance(rf_medium, "RF Medium (500k)")
imp_large  <- get_importance(rf_large,  "RF Large (1M)")

importance_all <- bind_rows(imp_small, imp_medium, imp_large)

kable(
  head(importance_all),
  caption = "Raw Variable Importance (First 6 Rows)"
)
```
# Normalize Importance Values
To compare importance accross models, I normalize each model's importance values by its maximum.
```{r}
importance_norm <- importance_all %>%
  group_by(model) %>%
  mutate(norm_importance = importance / max(importance)) %>%
  ungroup()

kable(
  head(importance_norm),
  caption = "Normalized Variable Importance (First 6 Rows)"
)
```
# Feature Importance Plots by Model
I create a function to plot normalized importance for a single model
```{r}
plot_importance <- function(df, model_name) {
  ggplot(
    df %>% filter(model == model_name),
    aes(x = reorder(feature, norm_importance),
        y = norm_importance)
  ) +
    geom_col(fill = "steelblue") +
    coord_flip() +
    labs(
      title = paste("Variable Importance —", model_name),
      x = "Feature",
      y = "Normalized Importance"
    ) +
    theme_minimal()
}

plot_importance(importance_norm, "RF Small (200k)")
plot_importance(importance_norm, "RF Medium (500k)")
plot_importance(importance_norm, "RF Large (1M)")
```
# Combined Importance Comparison
This plot compares feature importance accroos all three Random Forest models.
```{r}
ggplot(
  importance_norm,
  aes(x = feature,
      y = norm_importance,
      fill = model)
) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    title = "Variable Importance Comparison Across RF Models",
    x = "Feature",
    y = "Normalized Importance"
  ) +
  theme_minimal()
```
# Summary Table for Case Study
I reshape the normalized importance into a wide table , sorted by the largest model's importance.
```{r}
importance_summary <- importance_norm %>%
  select(feature, model, norm_importance) %>%
  tidyr::pivot_wider(
    names_from  = model,
    values_from = norm_importance
  ) %>%
  arrange(desc(`RF Large (1M)`))

kable(
  importance_summary,
  caption = "Feature Importance Summary Across Random Forest Models"
)
```
# Interpretation
The feature importance results provide a clear view of which variables most strongly influence the Random Forest models’ predictions of weekly Walmart sales. Across all three model sizes (200k, 500k, and 1M rows), several consistent patterns emerge.

1. Stable High‑Impact Features  
Certain features remain highly important regardless of sample size. These typically include variables tied to store characteristics, seasonality, and macroeconomic conditions. Their stability suggests they are fundamental drivers of Walmart’s sales patterns.

2. Increasing Clarity with Larger Samples  
As the training dataset grows, the importance rankings become more refined. The 1M‑row model shows sharper separation between top and mid‑tier features, indicating that more data helps the model better distinguish true signal from noise.

3. Holiday and Seasonal Effects  
Features related to holidays and temperature often rise in importance, reflecting the strong seasonal nature of retail demand. This aligns with known patterns such as holiday spikes and weather‑driven category shifts.

4. Economic Sensitivity  
Variables like fuel price and CPI show meaningful importance, suggesting that Walmart’s weekly sales respond to broader economic conditions. This is consistent with consumer behavior during inflationary or high‑fuel‑cost periods.

Overall, the feature importance analysis confirms that the model is capturing realistic, business‑aligned relationships between external factors, store attributes, and sales outcomes.

# Conclusion
This notebook provides a structured interpretation of the Random Forest models by analyzing and comparing feature importance across three dataset sizes. The results highlight:

Which features consistently drive sales predictions

How model understanding improves with more data

How external, seasonal, and economic factors shape Walmart’s sales patterns

These insights deepen the interpretability of the forecasting pipeline and support downstream decision‑making in pricing, promotions, inventory planning, and scenario analysis.

Notebook #4 bridges the gap between raw model performance (Notebook #3) and applied business forecasting (Notebook #5), completing the analytical story with clear, data‑driven explanations of why the model behaves the way it does.
