---
title: "04 — Feature Importance & Model Insights"
author: "Waldo Ketonou"
output: html_document
---
# Introduction
This notebook analyzes feature importance for the Random Forest models trained in the Walmart Sales Forecasting Project.
Models included:
RF Small (200k rows)
RF Medium (500k rows)
RF Large (1M rows)
Goals:
Identify which features most strongly drive weekly sales
Compare importance patterns across different sample sizes
Provide model interpretation for the final case study and dashboard
All data preparation and model training must be completed before running this notebook.
# Load Librarie
```{r}
library(dplyr)
library(tidyr)
library(ggplot2)
library(tibble)
library(knitr)
```
# Load Trained Models
This notebook assumes the Random Forest models have already been traine and saved as RDS in the outputs/models/ directory.
If you are running from a different working directory, adjust the paths accordingly.
```{r}
rf_small_path  <- "../outputs/models/rf_small.rds"
rf_medium_path <- "../outputs/models/rf_medium.rds"
rf_large_path  <- "../outputs/models/rf_large.rds"

rf_small  <- readRDS(rf_small_path)
rf_medium <- readRDS(rf_medium_path)
rf_large  <- readRDS(rf_large_path)
```
# Extract Variable Importance
I define a helper function to extract variable importance from each Random Forest model.
```{r}
get_importance <- function(model, model_name) {
  
  if (is.null(model$variable.importance)) {
    stop(paste("Variable importance not found for model:", model_name))
  }
  
  tibble(
    feature    = names(model$variable.importance),
    importance = as.numeric(model$variable.importance),
    model      = model_name
  )
}

imp_small  <- get_importance(rf_small,  "RF Small (200k)")
imp_medium <- get_importance(rf_medium, "RF Medium (500k)")
imp_large  <- get_importance(rf_large,  "RF Large (1M)")

importance_all <- bind_rows(imp_small, imp_medium, imp_large)

kable(
  head(importance_all),
  caption = "Raw Variable Importance (First 6 Rows)"
)
```
# Normalize Importance Values
To compare importance accross models, I normalize each model's importance values by its maximum.
```{r}
importance_norm <- importance_all %>%
  group_by(model) %>%
  mutate(norm_importance = importance / max(importance)) %>%
  ungroup()

kable(
  head(importance_norm),
  caption = "Normalized Variable Importance (First 6 Rows)"
)
```
# Feature Importance Plots by Model
I create a function to plot normalized importance for a single model
```{r}
plot_importance <- function(df, model_name) {
  ggplot(
    df %>% filter(model == model_name),
    aes(x = reorder(feature, norm_importance),
        y = norm_importance)
  ) +
    geom_col(fill = "steelblue") +
    coord_flip() +
    labs(
      title = paste("Variable Importance —", model_name),
      x = "Feature",
      y = "Normalized Importance"
    ) +
    theme_minimal()
}

plot_importance(importance_norm, "RF Small (200k)")
plot_importance(importance_norm, "RF Medium (500k)")
plot_importance(importance_norm, "RF Large (1M)")
```
# Combined Importance Comparison
This plot compares feature importance accroos all three Random Forest models.
```{r}
ggplot(
  importance_norm,
  aes(x = feature,
      y = norm_importance,
      fill = model)
) +
  geom_col(position = "dodge") +
  coord_flip() +
  labs(
    title = "Variable Importance Comparison Across RF Models",
    x = "Feature",
    y = "Normalized Importance"
  ) +
  theme_minimal()
```
# Summary Table for Case Study
I reshape the normalized importance into a wide table , sorted by the largest model's importance.
```{r}
importance_summary <- importance_norm %>%
  select(feature, model, norm_importance) %>%
  tidyr::pivot_wider(
    names_from  = model,
    values_from = norm_importance
  ) %>%
  arrange(desc(`RF Large (1M)`))

kable(
  importance_summary,
  caption = "Feature Importance Summary Across Random Forest Models"
)
```
