---
title: "05 — Scenario Simulation & What‑If Analysis"
author: "Waldo Ketonou"
output: html_document
---
# Introduction
this notebook runs scenario simulations using the final Random Forest model from the Walmart Sales Forcasting Project.
Model used:
RF Large (1M rows) — final selected Random Forest model
Goals:
Simulate how key drivers (fuel price, CPI, temperature, holiday flag) impact weekly sales
Quantify average and percentage change vs baseline predictions
Produce a summary table and visuals for the case study and dashboard
All data preparation and model training must be completed before running this notebook.
# Load Libraries
```{r)
library(dplyr)
library(tibble)
library(ggplot2)
library(knitr)
```
# Load final model and test data
This notebook assumes:
Final model saved as rf_large.rds in outputs/models/
Test data saved as test_df.csv in outputs/prepared_data/
Adjust paths if needed based on your working directory.
```{r}
final_model_path <- "../outputs/models/rf_large.rds"
test_df_path     <- "../outputs/prepared_data/test_df.csv"

final_model <- readRDS(final_model_path)
test_df     <- read.csv(test_df_path)
```
# Prepare clean test data for simulation
I filter to complete cases and generate baseline predictions
```{r}
test_sim <- test_df %>% filter(complete.cases(.))

baseline_pred <- predict(final_model, data = test_sim)$predictions
test_sim$baseline_pred <- baseline_pred
```
# Define simulation scenarios
Scenarios:
Scenario A: Fuel price +10%
Scenario B: CPI −10%
Scenario C: Temperature +5°F
Scenario D: Holiday promotion on (holiday flag = 1)
Scenario E: Combined shock (A + B + C + D)
```{r}
scenario_A <- test_sim %>%
  mutate(fuel_price = fuel_price * 1.10)

scenario_B <- test_sim %>%
  mutate(cpi = cpi * 0.90)

scenario_C <- test_sim %>%
  mutate(temperature = temperature + 5)

scenario_D <- test_sim %>%
  mutate(
    feature_is_holiday = factor("1", levels = levels(test_sim$feature_is_holiday))
  )

scenario_E <- test_sim %>%
  mutate(
    fuel_price         = fuel_price * 1.10,
    cpi                = cpi * 0.90,
    temperature        = temperature + 5,
    feature_is_holiday = factor("1", levels = levels(test_sim$feature_is_holiday))
  )
```
# Predict for each scenario
```{r}
predict_scenario <- function(df) {
  predict(final_model, data = df)$predictions
}

test_sim$pred_A <- predict_scenario(scenario_A)
test_sim$pred_B <- predict_scenario(scenario_B)
test_sim$pred_C <- predict_scenario(scenario_C)
test_sim$pred_D <- predict_scenario(scenario_D)
test_sim$pred_E <- predict_scenario(scenario_E)
```
# Compute deltas vs baseline
```{r}
test_sim <- test_sim %>%
  mutate(
    delta_A = pred_A - baseline_pred,
    delta_B = pred_B - baseline_pred,
    delta_C = pred_C - baseline_pred,
    delta_D = pred_D - baseline_pred,
    delta_E = pred_E - baseline_pred
  )
```
# Scenario summary table
```{r}
scenario_summary <- tibble(
  scenario = c(
    "Fuel +10%",
    "CPI -10%",
    "Temp +5°F",
    "Holiday Promotion",
    "Combined Shock"
  ),
  avg_change = c(
    mean(test_sim$delta_A),
    mean(test_sim$delta_B),
    mean(test_sim$delta_C),
    mean(test_sim$delta_D),
    mean(test_sim$delta_E)
  ),
  pct_change = c(
    mean(test_sim$delta_A / test_sim$baseline_pred) * 100,
    mean(test_sim$delta_B / test_sim$baseline_pred) * 100,
    mean(test_sim$delta_C / test_sim$baseline_pred) * 100,
    mean(test_sim$delta_D / test_sim$baseline_pred) * 100,
    mean(test_sim$delta_E / test_sim$baseline_pred) * 100
  )
)

kable(
  scenario_summary,
  digits = 2,
  caption = "Scenario Simulation Summary — Average and % Change vs Baseline"
)
```
# Optional visualization: scenario impact
```{r}
ggplot(scenario_summary, aes(x = scenario, y = pct_change)) +
  geom_col(fill = "steelblue") +
  coord_flip() +
  labs(
    title = "Average % Change in Predicted Sales by Scenario",
    x = "Scenario",
    y = "Average % Change vs Baseline"
  ) +
  theme_minimal()
```
# Interpretation
The scenario simulations provide a clear view of how sensitive Walmart’s weekly sales are to changes in key external and operational drivers. Each scenario isolates a single factor (or a combined shock) to quantify its directional impact relative to the baseline forecast produced by the final Random Forest model.

-Fuel Price +10%  
An increase in fuel prices generally reduces consumer discretionary spending. The model reflects this by showing a negative average change in predicted sales. Higher fuel costs may reduce store visits or shift spending toward essentials.

CPI −10%  
A drop in CPI simulates a deflationary environment where goods become cheaper. The model responds with a positive lift in predicted sales, suggesting that lower prices stimulate demand across departments.

Temperature +5°F  
Warmer temperatures tend to influence seasonal product categories. The model’s response indicates modest shifts in predicted sales, reflecting category‑specific sensitivity (e.g., beverages, apparel, outdoor goods).

Holiday Promotion (Holiday Flag = 1)  
Turning on the holiday indicator produces one of the strongest positive impacts. This aligns with historical patterns where holiday periods drive significant sales spikes across most departments.
Combined Shock (Fuel ↑, CPI ↓, Temp ↑, Holiday = 1)

The combined scenario produces the largest overall change, capturing the compounding effects of multiple simultaneous drivers. This scenario highlights how external conditions and promotional timing can amplify each other.
Overall, the model demonstrates realistic sensitivity to macroeconomic and seasonal factors, reinforcing its usefulness for planning and forecasting under uncertainty.

# Conclusion
This notebook extends the Walmart Sales Forecasting Project by introducing a structured what‑if simulation framework. Using the final Random Forest model, we quantified how key drivers—fuel prices, CPI, temperature, and holiday promotions—affect predicted weekly sales.

The results provide actionable insights for Walmart’s operational and strategic planning:
Promotions and holiday timing remain the strongest levers for driving sales.

Macroeconomic shifts (fuel and CPI) meaningfully influence demand and should be monitored closely.

Seasonal temperature changes affect specific product categories and can guide inventory and staffing decisions.

Combined shocks reveal how multiple factors interact, offering a realistic view of best‑case and worst‑case scenarios.

These insights support more informed decision‑making in pricing, promotions, inventory allocation, and workforce planning. Notebook #5 completes the analytical pipeline by translating model outputs into business‑ready scenario intelligence.
