---
title: "05 ŌĆö Scenario Simulation & WhatŌĆæIf Analysis"
author: "Waldo Ketonou | WaldoSphere Group LLC"
output: html_document
---

# Walmart Weekly Sales Forecasting  
## Notebook 05 ŌĆö Scenario Simulation & WhatŌĆæIf Analysis  

### Introduction  
This notebook performs businessŌĆæfocused scenario simulations using the **final Random Forest Large model (1M rows)**.  
It mirrors the production script `10_simulation_scenarios.R` and evaluates how predicted weekly sales respond to:

- Fuel price increase (+10%)  
- CPI decrease (ŌĆō10%)  
- Temperature increase (+5┬░F)  
- Holiday promotion activation  
- Combined shock scenario (all changes applied together)

These simulations quantify operational sensitivity and support the final case study and dashboard.

---

# 1. Load Libraries

```{r setup, message=FALSE, warning=FALSE}
library(dplyr)
library(readr)
library(ranger)

cat("\n===== Scenario Simulation: RF Large (1M rows) =====\n")
```
# 2. Load Final Model and Test Data
```{r}
final_model <- readRDS("outputs/models/rf_large_model.rds")

test_df <- read_csv("outputs/prepared_data/test_df.csv")
test_clean <- test_df %>% filter(complete.cases(.))
```
# 3. Baseline Prediction
We generate baseline predictions to compare against scenario outcomes.
```{r}
baseline_pred <- predict(final_model, data = test_clean)$predictions
test_clean$baseline_pred <- baseline_pred
```
# 4. Define Simulation Scenarios
Each scenario modifies one or more key drivers.
```{r}
scenario_A <- test_clean %>% mutate(fuel_price = fuel_price * 1.10)
scenario_B <- test_clean %>% mutate(cpi = cpi * 0.90)
scenario_C <- test_clean %>% mutate(temperature = temperature + 5)

scenario_D <- test_clean %>% mutate(
  feature_is_holiday = factor("1", levels = levels(test_clean$feature_is_holiday))
)

scenario_E <- test_clean %>% mutate(
  fuel_price         = fuel_price * 1.10,
  cpi                = cpi * 0.90,
  temperature        = temperature + 5,
  feature_is_holiday = factor("1", levels = levels(test_clean$feature_is_holiday))
)
```
# 5. Predict for Each Scenario
We use a helper function for clean, consistent predictions.
```{r}
predict_scenario <- function(df) {
  predict(final_model, data = df)$predictions
}

test_clean$pred_A <- predict_scenario(scenario_A)
test_clean$pred_B <- predict_scenario(scenario_B)
test_clean$pred_C <- predict_scenario(scenario_C)
test_clean$pred_D <- predict_scenario(scenario_D)
test_clean$pred_E <- predict_scenario(scenario_E)
```
# 6. Compute Deltas vs Baseline
We quantify the directional impact of each scenario.
```{r}
test_clean <- test_clean %>%
  mutate(
    delta_A = pred_A - baseline_pred,
    delta_B = pred_B - baseline_pred,
    delta_C = pred_C - baseline_pred,
    delta_D = pred_D - baseline_pred,
    delta_E = pred_E - baseline_pred
  )
```
# . Summary Table for Case Study
This table is used directly in the final report and dashboard.
```{r}
scenario_summary <- tibble(
  scenario = c(
    "Fuel +10%",
    "CPI -10%",
    "Temp +5┬░F",
    "Holiday Promotion",
    "Combined Shock"
  ),
  avg_change = c(
    mean(test_clean$delta_A),
    mean(test_clean$delta_B),
    mean(test_clean$delta_C),
    mean(test_clean$delta_D),
    mean(test_clean$delta_E)
  ),
  pct_change = c(
    mean(test_clean$delta_A / test_clean$baseline_pred) * 100,
    mean(test_clean$delta_B / test_clean$baseline_pred) * 100,
    mean(test_clean$delta_C / test_clean$baseline_pred) * 100,
    mean(test_clean$delta_D / test_clean$baseline_pred) * 100,
    mean(test_clean$delta_E / test_clean$baseline_pred) * 100
  )
)

scenario_summary
```
# 8. Save Summary to Disk
```{r}
write_csv(scenario_summary, "outputs/evaluation/rf_large_scenario_summary.csv")

cat("\nScenario simulation complete. Summary saved to /outputs/evaluation/\n")
```
# 9. Conclusion
This notebook successfully:

Applied five businessŌĆærelevant simulation scenarios

Generated predictions using the final RF Large model

Computed average and percentage changes vs baseline

Exported a clean summary table for the case study
